---
title: "memo-analysis"
output: html_document
---

```{r}

#importing libraries

library(tidyverse)
library(janitor)
library(readxl)
library(stringr)

```

```{r}

# loading in salary data

umd_salary_25 <- read_excel("data/Employee Annual Salaried-Diamondback 2025.xlsx") |> clean_names()

umd_salary_24 <- read_csv("data/diamondback-annual-salary-data-2024.csv") |> clean_names()

```

```{r}

# cleaning the salary from character with $ in front to numerical so we can add later
umd_salary_24 <-
umd_salary_24 |>
  mutate(
    appt_base_annual_salary = gsub("\\$", "", appt_base_annual_salary),
    appt_base_annual_salary = as.numeric(gsub(",","", appt_base_annual_salary)),
  year = 2024 #adding a year indicator
  )

# adding a year column to the 2025 dataset

umd_salary_25 <-
umd_salary_25 |>
  mutate(
    year = 2025
  )
```

```{r}

# Cleaning umd 2025 and 2024 salary to determine division breakdown. Beforehand the salary divisions were locked in a string that made it too un-useable


# first step is taking out the acronyms used the unit column and putting these pulled out strings into a new column named division_new

umd_salary_25 <-
umd_salary_25 |>
  mutate(
    division_new = gsub("[–—-].*$", "", unit) #replacing everything after the first dash with a bank, essentially removing everything after the first division reference
  )

# doing the same thing for the 2024 data set, its just located in a different column
umd_salary_24 <-
umd_salary_24 |>
  mutate(
    division_new = gsub("[–—-].*$", "", department)
  )

```

```{r}

# checking the contents of the new divisions column

umd_salary_25 |>
  group_by(division_new) |>
  summarise(
    count = n()
  )

# there's a bit of a white space issue that seems to be making somethings not group together properly

# fixing white space issue

umd_salary_25 <-
umd_salary_25 |>
  mutate(division_new = str_to_upper(division_new)) |> # chaning the new division string to all uppercase 
  mutate(division_new = str_trim(division_new)) # str_trim removes whitespace from the begining and end of a string

```

```{r}

# checking again

umd_salary_25 |>
  group_by(division_new) |>
  summarise(
    count = n()
  )

# there's a case where ENG is listed as the division though 

# investigating the rows where division is listed as ENG
umd_salary_25 |>
  filter(
    division == "ENG"
  )

# seems to be ENG happened only after I stripped out division from the unit but previously the division was SVPAAP so I'm just going to change it back for these two cases

umd_salary_25 <- 
umd_salary_25 |>
mutate(
    division_new = ifelse(division_new == "ENG", "SVPAAP", division_new)
  )

```

```{r}
# checking again

umd_salary_25 |>
  group_by(division_new) |>
  summarise(
    count = n()
  )

# there's divisions listed as VP FOR BELONGING and VPA PUBLIC SAFETY but there such small numbers that I'm going to check them

# inspecting VP FOR BELONGING (8 instances) and VPA PUBLIC SAFETY (3 instances)

umd_salary_25 |>
  filter(division_new == "VP FOR BELONGING" | division_new == "VPA PUBLIC SAFETY") |> 
  arrange(unit)

# the division_old lists them as either PRES of VPA ...  PRES for VP for Belonging and VPA for VPA Public Safety
# changing the division to match these cases, this will be fine because we will still know the previous unit name

umd_salary_25 <-
umd_salary_25 |>
  mutate(
    division_new = ifelse(division_new == "VP FOR BELONGING", "PRES", division_new)
  ) |>
  mutate(
    division_new = ifelse(division_new == "VPA PUBLIC SAFETY", "VPA", division_new)
  )
```

```{r}

# checking again

umd_salary_25 |>
  group_by(division_new) |>
  summarise(
    count = n()
  )

# looks like the only ones left is "PEITZMEIER FACULTY" since there was only 1 reference to it in the new data set

umd_salary_25 |>
  filter(division_new == "PEITZMEIER FACULTY")

# was previously labeled as SVPAAP, chaning division new to match this

umd_salary_25 <-
umd_salary_25 |>
  mutate(
    division_new = ifelse(division_new == "PEITZMEIER FACULTY", "SVPAAP", division_new)
  )

```

```{r}

# checking

umd_salary_25 |>
  group_by(division_new) |>
  summarise(
    count = n()
  )

# divisions seem all good now for 2025!

```

Before I calculate any median salaries I need to combine people with multiple rows

```{r}
# aggregate roles to combine multiple roles for one person
umd_salary_agg_25 <-
umd_salary_25 |>
  group_by(primary_name, unit) |> # grouping by department name and person name 
  summarise(
    appt_base_annual_salary = sum(appt_base_annual_salary, na.rm = TRUE),
    title = paste(unique(title), collapse = " & "),
    division_new = division_new,
    .groups = "drop"
  ) |>
  ungroup()
  
umd_salary_agg_25 |> # checking unique names 
group_by(primary_name, division_new) |> 
  filter(n() > 1) |>
  ungroup() |>
  arrange(primary_name)

# alright 18 people with the same name in the same division
# checking by unit

umd_salary_agg_25 |> # checking unique names 
group_by(primary_name, unit) |> 
  filter(n() > 1) |>
  ungroup() |>
  arrange(primary_name)

# okay great, only four instances where there is the same name and unit
# these four seem to be two instances of the same person since they are identical rows exactly
# I'm going to drop them to not skew median salary later

# just getting the pre-changed data set count of rows before and after to make sure it only cut the two that I wanted
nrow(umd_salary_agg_25) #10963

umd_salary_agg_25 <-
umd_salary_agg_25 |>
  group_by(primary_name, unit) |>
  unique() |> # now 10961, two less rows yay!
  ungroup()
  
umd_salary_agg_25 |>
group_by(primary_name, unit) |> 
  filter(n() > 1) |>
  ungroup() |>
  arrange(primary_name)
# beautiful there are now no duplicate rows

```

!!!!I WILL BE USING umd_salary_agg_25 FOR ALL ANALAYIS GOING FORWARD, unless noted otherwise

*University of Maryland president Darryll Pine's Salary increased to more than $1 million for the first time in 2025.*

*Pine's salary increased about 7 percent from more than $963,000 in 2024.*

```{r}

# agg data set not needed since just looking at one person

# getting 2024 salary for Pines

umd_salary_24 |> # using the 24 dataset
  filter(
    primary_name == "Pines, Darryll J." # filtering for just Pines
  ) |>
  select(year, primary_name, appt_base_annual_salary) # getting his name, the year of the salary and the salary ammount.

# getting 2025 salary for Pines

umd_salary_25 |> # using the 25 dataset
  filter(
    primary_name == "Darryll Pines" # filtering for only Pines
  ) |>
  select(year, primary_name, appt_base_annual_salary) # getting only the info we need and not including other columns

# 1032477 in 2025
# 963572 in 2024

# percent diff
# ((new value - old value) / old value) * 100
((1032477 - 963572) / 963572) * 100

# About a 7 percent increase to more than 1 million 

```

*Pines is now one of four university employees' whose salary is above $1 million, in 2024 only Maryland Men's Basketball head coach Mark Turgen had a salary of more than a million.*

```{r}

# not using agg data set since he's one employee

# filtering for university employees that had a salary of a million dollars or more in 2024
umd_salary_24 |>
  filter(
    appt_base_annual_salary >= 1000000
  )
# none except Mark Turgen ... Maryland Men's Basketball ex head coach 

# filtering for university employees that had a salary of a million dollars or more in 2025
umd_salary_25 |>
  filter(
    appt_base_annual_salary >= 1000000
  )
# 4 people had a salary of a million or more including Turgen ... Pines, Jim Smith and Brian Williams. Brian Williams does not currently work at this university.

# so it is just Pines and Smith who have a salary of more than a million

```



*Appointed in May of 2025, university director of athletics James E. Smith’s salary is currently $1.2 million, the highest out of all employees currently working at the university according to the paper’s analysis.*

Note: Since Mark Turgeon is no longer with the university he is not counted as a current employee and is then not the top employee salary at this university in 2025.

```{r}

# not using agg since just one person

umd_salary_25 |> # using 25 data
  arrange(desc(appt_base_annual_salary)) |> # arranging from highest salaries to lowest salaries
  select(primary_name, appt_base_annual_salary) # only asking for the same of the employee and their salary, other info not needed

```

*Damon Evan, the previous director of athletics for seven years, had a salary of more than $836,000 in 2024 ...*

```{r}

# not using agg because one person

umd_salary_24 |> # 2024 data set
  filter(primary_name == "Evans, Damon") |> # getting only Damon Evans
  select(primary_name, title, appt_base_annual_salary)

```

*about 44 percent less than Smith’s first year salary at this university.*

for section head as well
*James E. Smith’s salary is more than $363,000 higher than the last director of athletics.*

```{r}

# not using agg because comparing just two individual people

# Smith's salary in 2025: 1200000
# Evans salary in 2024: 836256.3

1200000 - 836256.3

# difference of $363,743.7 between Smith and Evans

((1200000 - 836256.3) / 836256.3) * 100

# percent difference 43.49668 increase from Evans to Smith's salary

```

*The athletic department makes up a majority of the highest-paid employees at this university.*

```{r}

umd_salary_agg_25 |> # 25 data
  arrange(desc(appt_base_annual_salary)) |> # highest salary to least
  select(unit, appt_base_annual_salary, title) |> # only unit, salary and title
  head(10) # only the employees with the top 10 salaries

# of the 10 most paid employees at the university 7 are  within the ICA/Intercollegiate Athletics unit. 

# Note: this includes those not currently with the department notably Turgen and ex MD football defensive coordinator Brian Williams 

```

To find the median salary of employees within the intercollegiate athletics division I needed to separate all employees from the division from the larger data set.

```{r}

# found that there were the following unit categorizations for those involved in the intercollegiate athletic department

# used this do find it since I realized that they were all located at least within the President divison:

umd_salary_agg_25 |>
  filter(str_detect(unit, "PRES") | str_detect(unit, "ICA")) |>
  group_by(unit) |>
  summarise(
    count = n()
  )

# PRES-ICA-Director's Office/Admin	
# PRES-ICA-External Relations	
# PRES-ICA-Football	
# PRES-ICA-Men's Basketball
# PRES-ICA-Operations	
# PRES-ICA-Other Men's Sports	
# PRES-ICA-Sports Admin (Overloads/Nonstandards)	
# PRES-ICA-Support Services	
# PRES-ICA-Women's Sports	
# PRES-Intercollegiate Athletics

# creating a filter so that I can find the median salary for those within the department / a count for the number employed in this department


ica_25_filter <-
umd_salary_agg_25 |>
  filter(
    unit == "PRES-ICA-Director's Office/Admin" |
    unit == "PRES-ICA-External Relations" |
    unit == "PRES-ICA-Football" |
    unit == "PRES-ICA-Men's Basketball" |
    unit == "PRES-ICA-Operations" |
    unit == "PRES-ICA-Other Men's Sports" |
    unit == "PRES-ICA-Sports Admin (Overloads/Nonstandards)" |
    unit == "PRES-ICA-Support Services" |
    unit == "PRES-ICA-Women's Sports" |
    unit == "PRES-Intercollegiate Athletics" 
  )

```

*The median salary for an employee at the intercollegiate athletic department makes just under $100,000*

```{r}

ica_25_filter |> # using the previously created filter
  summarise(
    median = median(appt_base_annual_salary) # finding median salary
  )

# median salary for those within the athletic department is $99,545.77

```

*Despite his absence from Maryland since 2021, former Maryland men’s basketball head coach Mark Turgen’s buyout deal makes him the highest-paid person at this university.*

```{r}

umd_salary_25 |> # 25 data set
  arrange(desc(appt_base_annual_salary)) |> # salary from most to least
  head(1) |> # getting only the top base annual salary
  select(title, primary_name, appt_base_annual_salary)

```

*Former Maryland football defensive coordinator Biran Williams is the third most-paid employee since his departure in the spring of 2025, according to 2025 data.*

```{r}

umd_salary_25 |> # 25 data set 
  arrange(desc(appt_base_annual_salary)) |> # salary most to least 
  head(3) |> # top 3 most paid
  select(
    primary_name, title, appt_base_annual_salary
  )

```

*Turgen’s salary was $1.25 million and Williams’ salary was reported at $1.06 million, according to 2025 salary data.*

```{r}

# Mark Turgeon's salary
umd_salary_25 |>
  filter(primary_name == "Mark Turgeon") |>
  select(primary_name, appt_base_annual_salary, title)

# Williams salary
umd_salary_25 |>
  filter(primary_name == "Brian Williams") |>
  select(primary_name, appt_base_annual_salary, title)

```

*Turgen’s buyout is more than current men’s basketball head coach Brent “Buzz” Williams’ $500,000 salary.*

```{r}

# Buzz Williams' salary

umd_salary_25 |>
  filter(primary_name == "Brent Williams") |>
  select(primary_name, appt_base_annual_salary, title)


```

```{r}

# Next step is to filter for only academic departments at the university

# Creating a filter for the 2025 data for only academic departments
# Filtering for what we don't want
salary_25_academic <-
umd_salary_agg_25 |> # filtering out everything that isn't an academic department
                 # decided manually by just looking up each acronym on google and checking what they represented
  filter(!(division_new == "DIT" | division_new == "VPR" | division_new == "VPUR" | division_new == "GRAD" | division_new == "PRES" | division_new == "SVPAAP" | division_new == "USG" | division_new == "UGST" | division_new == "LIBR" | division_new == "VPA" | division_new == "VPSA" | division_new == "EXST" | division_new == "VPF")) |>
  mutate(
    title = str_to_lower(title)
  )

```

*The school of business’ median employee salary ranked highest among academic divisions*

*The median salary for an employee at the Robert H. Smith school of business was more than $135,000*

```{r}

salary_25_academic |>
group_by(division_new) |>
  summarise(
  sumemployees = n(),
  median_salary = median(appt_base_annual_salary),
) |>
  arrange(desc(median_salary))

```

*at least 27 percent higher than any other academic division at this university in 2025*

```{r}

# finding % diff between BMGT mean salary BSOS (the second highest mean employee salary)

((135055.34 - 105778.18) / 105778.18) *100

```

*Of the 12 academic divisions at this university, seven had a median employee salary of more than $100,000 during 2025.*

*The college of agriculture and natural resources had a median employee salary of under $80,000, the lowest among academic departments.*

```{r}

salary_25_academic |>
group_by(division_new) |>
  summarise(
  sumemployees = n(),
  median_salary = median(appt_base_annual_salary),
) |>
  arrange(desc(median_salary))

```

*Among academic divisions, individuals from the business school held four of the top five individual salaries. 64 percent of business school employees have a median salary more than $100,000, the highest ratio among other academic divisions.*

```{r}

salary_25_academic |>
group_by(division_new) |>
  summarise(
  n_employees = n(),
  salary_morethan_100000 = sum(appt_base_annual_salary > 100000),
  pct_morethan100000 = (salary_morethan_100000 / n_employees) *100
) |>
  arrange(desc(pct_morethan100000))

```

